---
title: "589Project"
author: "Yuyang Peng"
date: "2024-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data process
### read xlsx file

```{r}
# Install and load the readxl package if not already installed
if (!requireNamespace("readxl", quietly = TRUE)) {
  install.packages("readxl")
}
library(readxl)
library(dplyr)

# Read the Excel file and convert it to a dataframe
df <- read_excel("Hornets data.xlsx")


# Filter the dataframe and remove rows with missing longitude or latitude
target_df <- df %>%
  filter(stateProvince %in% c('British Columbia', 'Canada - British Columbia (BC)')) %>%
  filter(!is.na(decimalLongitude) | !is.na(decimalLatitude))
```

### Load BC_Covariates

```{r}
if (!requireNamespace("maptools", quietly = TRUE)) {
  install.packages("maptools", repos = "https://packagemanager.posit.co/cran/2023-10-13")
}
# Load required packages
library(sp)
library(spatstat)
library(sf)
library(maptools)
# Import BC_Covariates
load("BC_Covariates.Rda")
window = DATA$Window
```

### Plot distribution

```{r}
library(ggplot2)

# Plot longitude and latitude coordinates
ggplot(target_df, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_point(color = "blue") +  # Add points
  labs(x = "Longitude", y = "Latitude",  # Label axes
       title = "Location Coordinates for Hornet within BC",  # Add title
       color = "Data Points") +  # Legend title
  scale_color_manual(values = "blue", name = "Data Points") +  # Legend color and title
  theme_minimal() +  # Apply minimal theme
  theme(
    panel.grid = element_blank(),  # Remove grid lines
    axis.line = element_line(color = "black")  # Color axis lines
  )

```

### Projection process

```{r}
library(sp)

# Define the projection string
proj_string <- "+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000 +y_0=0 +datum=NAD83 +units=m +no_defs"

# Create a SpatialPoints object from latitude and longitude
points <- SpatialPoints(coords = target_df[, c("decimalLongitude", "decimalLatitude")],
                        proj4string = CRS("+proj=longlat +datum=WGS84"))

# Transform the points to BC Albers projection
points_bc_albers <- spTransform(points, CRS(proj_string))

# Extract X and Y coordinates
X <- coordinates(points_bc_albers)[, 1]
Y <- coordinates(points_bc_albers)[, 2]

# Create a dataframe with X and Y columns
df_bc_albers <- data.frame(X = X, Y = Y)

# Now, 'df_bc_albers' contains the X and Y columns in the BC Albers projection

```

### Convert to ppp object

```{r}
# remove outside points
valid_points <- df_bc_albers[inside.owin(df_bc_albers$X, df_bc_albers$Y, as.owin(DATA$Window)), ]
hornet_ppp <- ppp(
  x = valid_points$X,  # X coordinates of valid points
  y = valid_points$Y,  # Y coordinates of valid points
  window = as.owin(DATA$Window)  # Observation window
)
plot(hornet_ppp)
```

## Exploratory Data Analysis
### Intensity calculation
From the plot, we can see most hornets tend to gather in lower BC, showing high intensity especially at the southwest corner. But for other part, especially for upper BC, there is no evidence of correlation in hornet location from eyes, indicating the distribution is inhomogeneous.

```{r}
win_km <- rescale(Window(hornet_ppp), 1000, "km")

# Intensity in trees/km^2
npoints(hornet_ppp)/area(win_km)
```

I dont think the estimated intensity trustworthy. Because from the plot we cannot tell the hornet location is homogeneous.

### quadrat tets

```{r}
Q <- quadratcount(hornet_ppp,
                  nx = 10,
                  ny = 10)
quadrat.test(Q)
```

The small p-value suggests that there is a significant deviation from homogeneity.

```{r}
plot(Q, main = "Quadrat Plot with Hornet Points", xlab = "X Coordinate", ylab = "Y Coordinate")
points(hornet_ppp, pch = 20, col = "green")  # Overlay hornet points
```

```{r}
intensity(Q)
plot ( intensity ( Q , image = T))
```
```{r}
lambda_u <- mean (Q)
R_Poisson <- rpois (n = length (Q) ,
lambda = lambda_u)
hist(Q)
hist(R_Poisson)
```
These data are clearly inhomogeneous.

### Kernel estimation with likelihood cross validation bandwidth selection & Hotspot analysis

```{r}
plot(density(hornet_ppp, sigma = bw.ppl),  # Likelihood Cross Validation Bandwidth Selection
     ribbon = F,
     main = "Hornet Kernel Estimation")
```

```{r}

R <- bw.ppl(hornet_ppp)

#Calculate test statistic
LR <- scanLRTS(hornet_ppp, r = R)

#Plot the output 
plot(LR,main='Hotspot Analysis')
plot(window,add = TRUE)
```

### Ripley's $K$-function, test for a significant ( $\alpha$ = 0.05) correlation between hornet locations

```{r}
#Estimate a strictly positive density
lambda_hornet_pos <- density(hornet_ppp,
                          sigma=bw.ppl,
                          positive=TRUE)

#Simulation envelope (with points drawn from the estimated intensity)
E_hornet_inhom <- envelope(hornet_ppp,
                        Kinhom,
                        simulate = expression(rpoispp(lambda_hornet_pos)),
                        correction="border",
                        rank = 1,
                        nsim = 19,
                        fix.n = TRUE)
```

```{r}
# visualise the results
par(mfrow = c(1,2))
plot(E_hornet_inhom,
     main = "",
     lwd = 2)
# Zoom in on range where significant deviations appear
plot(E_hornet_inhom,
     xlim = c(0,30000),
     main = "",
     lwd = 2)
```

When corrected for inhomogeneity, significant clustering only appears to exist in and around 0-22000 meters. For longer distance, we cannot indicate a significant evidence of correlation in hornet distribution.

### Pair correlation function (95% confidence interval)

```{r}
pcf_hornet_inhom <- envelope(hornet_ppp,
                          pcfinhom,
                          simulate = expression(rpoispp(lambda_hornet_pos)),
                          rank = 1,
                          nsim = 19)
par(mfrow = c(1,3))
plot(pcf_hornet_inhom)

# Zoom in on range where significant deviations appear
plot(pcf_hornet_inhom,
     xlim = c(0,10000),
     main = "",
     lwd = 2)

plot(pcf_hornet_inhom,
     xlim = c(15000,50000),
     main = "",
     lwd = 2)
```

There appear to be more hornet than expected by random chance between ∼ 0 - 6000 meters and less hornet between 15000 - 50000 meters. Except that, the locations of hornet appear not to exhibit any significant correlations.

## Single Indicator Exploring
### Distribution and Elevation

```{r}
plot(DATA$Elevation, main = "Elevation Map of British Columbia", xlab = "Longitude", ylab = "Latitude")

# Plot park locations on top of elevation map
plot(hornet_ppp, add = TRUE,cex = 0.4)

```

```{r}
# Define elevation classes based on quantiles
elevation_classes <- cut(DATA$Elevation, breaks = 5, labels = FALSE)

# Plot the elevation class image
image(elevation_classes, col = terrain.colors(5), main = "Elevation Classes")
plot(hornet_ppp, add = TRUE,cex = 0.4)
```

From plot, we could find that most hornets gather in Class 1 and 2.

```{r}
# Median elevation within British Columbia
median_bc_elevation <- median(DATA$Elevation, na.rm = TRUE)

# Median elevation at park locations
median_hornet_elevation <- median(DATA$Elevation[hornet_ppp], na.rm = TRUE)

# Print results
print(paste("Median elevation within British Columbia:", median_bc_elevation))
print(paste("Median elevation at hornet locations:", median_hornet_elevation))
```

```{r}
BC_elevation_density <- density(as.data.frame(DATA$Elevation, na.rm = TRUE)$value)

# Extract elevation values from DATA$Elevation using hornet_ppp
elevation_values <- as.vector(DATA$Elevation[hornet_ppp])

# Generate a kernel density estimate of the distribution of elevation values within hornet
hornet_elevation_density <- density(elevation_values, na.rm = TRUE)

plot(BC_elevation_density, col = "blue", main = "Kernel Density Estimate of Elevation in British Columbia", xlab = "Elevation",ylim = c(0, max(hornet_elevation_density$y)))

# Overlay the kernel density estimate of elevation values for park locations
lines(hornet_elevation_density, col = "red")

# Add legend
legend("topright", legend = c("British Columbia", "hornet Locations"), col = c("blue", "red"), lty = 1, cex = 0.8)
```


## Forest and Distribution

```{r}
plot(DATA$Forest, main = "Forest Map of British Columbia", xlab = "Longitude", ylab = "Latitude")

# Plot park locations on top of elevation map
plot(hornet_ppp, add = TRUE,cex = 0.4)
```
```{r}
# Define elevation classes based on quantiles
forest_classes <- cut(DATA$Forest, breaks = 5, labels = FALSE)

# Plot the elevation class image
image(forest_classes, col = terrain.colors(5), main = "Forest Classes")
plot(hornet_ppp, add = TRUE,cex = 0.4)
```
From plot, we could find that most hornets gather in Class 4 and 5.

```{r}
# Median forestdensity within British Columbia
median_bc_forest <- median(DATA$Forest, na.rm = TRUE)

# Median forest density at park locations
median_hornet_forest <- median(DATA$Forest[hornet_ppp], na.rm = TRUE)

# Print results
print(paste("Median Forest Density within British Columbia:", median_bc_forest))
print(paste("Median Forest Density at hornet locations:", median_hornet_forest))
```

```{r}
BC_forest_density <- density(as.data.frame(DATA$Forest, na.rm = TRUE)$value)

# Extract forest values from DATA$Forest using hornet_ppp
forest_values <- as.vector(DATA$Forest[hornet_ppp])

# Generate a kernel density estimate of the distribution of forest values within hornet
hornet_forest_density <- density(forest_values, na.rm = TRUE)

plot(BC_forest_density, col = "blue", main = "Kernel Density Estimate of Forest in British Columbia", xlab = "Forest",ylim = c(0, max(BC_forest_density$y)))

# Overlay the kernel density estimate of forest values for hornet locations
lines(hornet_forest_density, col = "red")

# Add legend
legend("topright", legend = c("British Columbia", "hornet Locations"), col = c("blue", "red"), lty = 1, cex = 0.8)
```
### Distance to Water Map

```{r}
plot(DATA$Dist_Water, main = "Distance to Water Map of British Columbia", xlab = "Longitude", ylab = "Latitude")

# Plot park locations on top of elevation map
plot(hornet_ppp, add = TRUE,cex = 0.4,col = 'white')
```
From the graph we can find that the distance is almost pretty low from water within BC, so i tend not to go deeper into the single indicator.

### HFI
```{r}
plot(DATA$HFI, main = "HFI Map of British Columbia", xlab = "Longitude", ylab = "Latitude")

# Plot park locations on top of elevation map
plot(hornet_ppp, add = TRUE,cex = 0.4, col='white')
```
From the graph we can find that HFI is almost pretty low within BC, so i tend not to go deeper into the single indicator.

## Poisson Point Process for Indicators
### Model Fitting
### Estimate $\rho$ for the locations of parks as a function of elevation/forest/water distance/HFI
```{r}
elev <- DATA$Elevation
hornet_elev_rho <- rhohat(hornet_ppp, elev)
plot(hornet_elev_rho,xlim = c(0,max(elev)))
```

```{r}
fors <- DATA$Forest
hornet_forest_rho <- rhohat(hornet_ppp, fors)
```

```{r}
plot(hornet_forest_rho,xlim = c(0,max(fors)))
```
```{r}
dist_water <- DATA$Dist_Water
```

```{r}
hornet_dist_rho <- rhohat(hornet_ppp, dist_water)
```

```{r}

plot(hornet_dist_rho,xlim = c(0,max(dist_water)))
```

```{r}
hfi <- DATA$HFI
```


So we could find that the three indicators cannot be linear with hornet distribution, which lead to the formula below.
```{r}
fit <- ppm(hornet_ppp ~ elev  + fors  + dist_water +  hfi , data = DATA)

fit

```
From the Ztest, we'd better remove forest

```{r}
fit_1 <- ppm(hornet_ppp ~ elev +  dist_water + hfi, data = DATA)

fit_1
```

```{r}
fit_2 <- ppm(hornet_ppp ~ elev + I(elev^2)+ dist_water + hfi + I(hfi^2), data = DATA)

fit_2
```

The coefficients are all statistically significant.

```{r}
#Plot the model predictions
plot(fit_2,
     se = FALSE,
     superimpose = FALSE)

#Overlay the B. pendula locations
plot(hornet_ppp,
     pch = 16,
     cex = 0.4,
     cols = "white",
     add = TRUE)
plot(hornet_ppp,
     pch = 16,
     cex = 0.3,
     cols = "black",
     add = TRUE)
```

### Model Selection
```{r}
AIC(fit_2);AIC(fit_1) 
```

```{r}
BIC(fit_2); BIC(fit_1)
```
We could find that AIC and BIC fall down when we remove one feature. So the extra complexity is not well supported by the data.

### Model Validation
```{r}
#Run the quadrat test
quadrat.test(fit_2, nx = 10, ny = 10)
```
The small p value tells us that there’s a significant deviation from our model’s predictions. While this is useful for suggesting that our model has room for improvement, it provides us with no direction on how to do so (e.g., missing parameters, model mispecification (e.g., polynomial vs. linear), a lack of independence, non-stationarity, etc…).

```{r}
#Calculate the residuals
res <- residuals(fit_2)

na_indexes <- which(is.na(res$val))
# If you need to exclude NA values explicitly
if (length(na_indexes) > 0) {
  res <- res[-na_indexes]
  plot(res, cols='transparent')
} else {
  plot(res, cols='transparent')
}
```
We can see some pattern in the residual plot. We still need to check

```{r}
#Calculate the partial residuals as a function of elevation
par_res_elev0 <- parres(fit_2, "elev")

#Calculate the relative intensity as a function of for
par_res_hfi0 <- parres(fit_2, "hfi")

par_res_water0 <- parres(fit_2,"dist_water")
#Side by side plotting
par(mfrow = c(1,3))
plot(par_res_elev0,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Elevation (m)")
plot(par_res_hfi0,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Forest")
plot(par_res_water0,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Distance to Water")
```
The plot shows that each feature could explain the intensity well.

```{r}
library(splines)

#Fit the PPP model
fit_smooth <- ppm(hornet_ppp ~ bs(elev,4) + bs(hfi, 7) + bs(dist_water,2), data = DATA, use.gam = TRUE)

fit_smooth
```
```{r}
coef(summary(fit_smooth))
```


```{r}
#Calculate the partial residuals as a function of elevation
par_res_elev <- parres(fit_smooth, "elev")

#Calculate the relative intensity as a function of for
par_res_hfi <- parres(fit_smooth, "hfi")

par_res_water <- parres(fit_smooth,"dist_water")
#Side by side plotting
par(mfrow = c(1,3))
plot(par_res_elev,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Elevation (m)")
plot(par_res_hfi,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "HFI")
plot(par_res_water,
     legend = FALSE,
     lwd = 2,
     main = "",
     xlab = "Distance to Water")
```
It is not good, but better than previous one.

```{r}
AIC(fit_2);AIC(fit_smooth)
```

```{r}
BIC(fit_2);BIC(fit_smooth)
```

```{r}
anova(fit_2, fit_smooth, test = "LRT")
```
The ANOVA results indicate that Model 2 has significantly better fit to the data than Model 1, as evidenced by the large difference in deviance (516.81) and the highly significant p-value (< 2.2e-16), which is well below any conventional significance level.
All lines of evidence point towards these more complex models being a better fit to the data. The final model is too long to write down, but we can visualise the predictions just as before.
```{r}
#Calculate the residuals
res1 <- residuals(fit_smooth)

na_indexes1 <- which(is.na(res1$val))
# If you need to exclude NA values explicitly
if (length(na_indexes1) > 0) {
  res1 <- res[-na_indexes1]
  plot(res1, cols='transparent')
} else {
  plot(res1, cols='transparent')
}
```

```{r}
#Plot the model predictions
plot(fit_smooth,
     se = FALSE,
     superimpose = FALSE,
     main = "Estimated Hornet intensity")

plot(hornet_ppp,
     pch = 16,
     cex = 0.6,
     cols = "white",
     add = TRUE)
plot(hornet_ppp,
     pch = 16,
     cex = 0.5,
     cols = "black",
     add = TRUE)
```

